---
title: "Computing and visualizing functional residuals with the funres package"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  echo = TRUE,
  cache = FALSE,
  fig.width = 6,
  fig.asp = 0.618,
  out.width = "70%"
)
```

```{r setup}
library(funres)
```


## Currently supported models

| Model type  | Family        | Package   | Function |
|:------------|:-------------|:-----------|:---------|
| GLM         | Binomial      | **stats** |  `glm()` |
| GLM         | Poisson       | **stats** |  `glm()` |
| GLM         | Quasi-Poisson | **stats** |  `glm()` |
| GAM         | Binomial      | **mgcv**  |  `gam()` |
| GAM         | Poisson       | **mgcv**  |  `gam()` |
| GAM         | Quasi-Poisson | **mgcv**  |  `gam()` |
| Ordinal     | NA            | **VGAM**  |  `vgam()` and `vgam()` |


## Logistic regression

We'll start by generating data from a quadratic logistic regression (LR) model:
```{r}
# Generate data from a logistic regression model with quadratic form
set.seed(1217)
n <- 1000
x <- rnorm(n)
# x[1] <- 10  # add an outlier
z <- 1 - 2*x + 3*x^2 + rlogis(n)
y <- ifelse(z > 0, 1, 0)

# Fit a couple of LR models
fit.bad <- glm(y ~ x, family = binomial)  # wrong
fit.good <- glm(y ~ x + I(x^2), family = binomial)  # right
```


### Functional residuals

```{r}
fres.bad <- fresiduals(fit.bad)
fres.good <- fresiduals(fit.good)
par(mfrow = c(1, 2), las = 1)
plot(fres.bad[[1]], xlab = "t", ylab = "F(t)")
plot(fres.good[[1]], xlab = "t", ylab = "F(t)")  # plot FRs for first observation
```

Does plotting them all tell us anything interesting?
```{r}
plot(fres.bad[[1]], xlab = "t", ylab = "F(t)", las = 1, type = "n")
tt <- 0:100/100
for (i in seq_along(fres.bad)) {
  lines(tt, fres.bad[[i]](tt), col = adjustcolor(1, alpha.f = 0.05))
}
```


### Surrogate and probability-scale residuals

```{r}
sr.bad <- fresiduals(fit.bad, type = "surrogate")
sr.good <- fresiduals(fit.good, type = "surrogate")
par(mfrow = c(1, 2))
col <- adjustcolor(1, alpha.f = 0.2)
plot(x, y = sr.bad, col = col, las = 1, ylab = "Surrogate residual")
lines(lowess(x, y = sr.bad), lwd = 2, col = "red2")
plot(x, y = sr.good, col = col, las = 1, ylab = "Surrogate residual")
lines(lowess(x, y = sr.good), lwd = 2, col = "red2")
```

```{r}
# Probability-scale residuals
ps.bad <- fresiduals(fit.bad, type = "probscale")
ps.good <- fresiduals(fit.good, type = "probscale")
par(mfrow = c(1, 2))
plot(x, y = ps.bad, col = col, las = 1, ylab = "Surrogate residual")
lines(lowess(x, y = sr.bad), lwd = 2, col = "red2")
plot(x, y = ps.good, col = col, las = 1, ylab = "Surrogate residual")
lines(lowess(x, y = sr.good), lwd = 2, col = "red2")
```



### Functional residual density (FRED) plots

The FRED plots in R are based on the Trellis framework (e.g., **lattice**), which rely on **grid** graphics. The `gridExtra::grid.arrange()` function is the most convenient approach to arranging several plots here. These graphs are quicker to produce compared to **ggplot2**.

```{r}
# Two-dimensional kernel density estimation
gridExtra::grid.arrange(
  fredplot(fit.bad, x = x),
  fredplot(fit.good, x = x),
  nrow = 1
)

# Hexagonal binning
gridExtra::grid.arrange(
  fredplot(fit.bad, x = x, type = "hex", aspect = 1),
  fredplot(fit.good, x = x, type = "hex", aspect = 1),
  nrow = 1
)
```


### Function-function plots

```{r}
palette("Okabe-Ito")
par(mfrow = c(1, 2))
ffplot(fit.bad, n = 500, type = "l")
ffplot(fit.good, n = 500, type = "l")
palette("default")
```


## Ordinal model example with VGAM package

```{r, out.width="100%"}
library(VGAM)

# Fit models to simulated ordinal response data
qdf <- sure::df1  # quadratic data from the sure package
fit1 <- vglm(y ~ x, data = qdf, family = acat(reverse = TRUE, parallel = TRUE))
fit2 <- vglm(y ~ poly(x, degree = 2), data = qdf, family = acat(reverse = TRUE, parallel = TRUE))
fit3 <- vgam(y ~ s(x), data = qdf, family = acat(reverse = TRUE, parallel = TRUE))

# Residual plots
gridExtra::grid.arrange(
  fredplot(fit1, x = qdf$x),  # linear term (wrong)
  fredplot(fit2, x = qdf$x),  # quadratic term
  fredplot(fit3, x = qdf$x),  # smooth term
  nrow = 1
)
```
